A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-19 tex/codatadp.tex
use a bracketing notation for normal forms, so we don't need an ugly combination of parentheses and arrow (if it were not for the parentheses, I would have liked the arrow)

%\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\vec}[1]{\bar #1}
%\newcommand{\Ec}{\mathsf{E}}
\newcommand{\Ec}{E}
%\newcommand{\Ec}{\mathcal{E}}
%\newcommand{\Fc}{\mathsf{F}}
\newcommand{\Fc}{F}
%\newcommand{\Fc}{\mathcal{F}}
%\newcommand{\Gc}{\mathsf{G}}
%\newcommand{\Ac}{\mathsf{A}}
%\newcommand{\Dc}{\mathsf{D}}
%\newcommand{\Rc}{\mathsf{R}}
\newcommand{\vrange}{\mathsf{range}}
\newcommand{\vdom}{\mathsf{dom}}
\newcommand{\tEc}{\mathcal{T}(\Ec)}
\newcommand{\tcEc}{\mathcal{T}^\ast(\Ec)}
\newcommand{\rn}[1]{\textsf{\small #1}}
\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\tneq}{\not\teq}
\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
%\newcommand{\is}[1]{is\text{-} #1}
\newcommand{\is}[1]{\const{is#1}}
\newcommand{\ror}{\kern.45em \parallel \kern.45em}
\newcommand{\tpath}[2]{\mathcal{P}_{ #2 \rightarrow \_ }( #1 )}
\newcommand{\ttpath}[3]{\mathcal{P}_{ #2 \rightarrow #3 }( #1 )}
\newcommand{\interp}[1]{\llbracket #1 \rrbracket}
\newcommand{\dpath}[3]{\delta^{#1}_{#2}( #3 )}
\newcommand{\ec}[1]{[#1]}
\newcommand{\M}{\mathcal{M}}
\newcommand{\Val}{\mathcal{A}}
\newcommand{\Var}{\mathcal{V}}
%\newcommand{\Varec}[1]{\Var{\ec{#1}}}
\newcommand{\Varec}[1]{\widetilde{#1}}
\newcommand{\VAREC}[1]{\widetilde{\vphantom{\scriptstyle x^i}\smash{#1}}}
\newcommand{\nf}[1]{{{#1}{\downarrow}}}
\newcommand{\aequiv}{\mathrel{=_\alpha}}
\newcommand{\vsim}{\aequiv}
\newcommand{\vsimv}[1]{\mathrel{=^{#1}_\alpha}}
\newcommand{\tpos}[2]{#1\!\mid_{#2}}
\newcommand{\muvar}{\mathcal{Var}}

@@

For any $\mu$-term $t$ of the form $\MU x.\; \const C( \vec u )$, we can
recursively construct the \emph{normal form} $\nf{t}$ of $t$,
that is a renaming of $v$ by replacing all of the bisimilar subterms of $t$ with $x$
%and by replacing all of the nonnormal subterms of $t$ with their corresponding normal forms.
and by normalizing $t$'s subterms.
For example, $\nf{(\MU x.\; \const{C}( \MU y.\; \const{C}( x ) ))} = \MU x.\; \const{C}( x )$.

\begin{lemma}
\label{lem:mu-norm-arg}
If $\vec u$ are normal and $t = \nf{(\MU x.\; \const{C}( \vec u ) )} = \MU x.\; \const{C}( \vec w )$,
then $\vec u \vsim t\interp{\vec w}$.
\end{lemma}
\begin{proof}
%Since $u_i$ is normal, $w_i$ is the result of replacing all subterms $v$ of $u_i$ such that $t\interp{v} \aequiv t$ with $x$.
\rem{TODO}
\end{proof}

@@

\begin{definition}[Normal Model]
\afterDot%
\label{def:norm-model}%
\rm
A model $\M$ is \emph{normal} if and only if:
\begin{enumerate}
\item
for all types $\tau$,
$\M( \tau )$ is a maximal set of closed normal $\mu$-terms of that type that are
unique up to renaming of $\mu$-bound variables,
and acyclic when $\tau \not\in \Codata$,
\item
for all constructor terms $\const{C}( \vec t \,)$ of type $\tau$,
$\M( \const{C}( \vec t \,) )$ is the value
in $\M( \tau )$ that is $\alpha$-equivalent to
$\nf{(\MU x.\; \const{C}( \M( \vec t \,) ))}$ where $x$ is fresh, and
\item
for all selector terms $\const s^j_i( t )$ of type $\tau$,% for some $\vec u$,
$\M( \const s^j_i( t ) )$ is the value
in $\M( \tau )$ that is $\alpha$-equivalent to
$\M( t )\interp{u_j}$
when $\M( t )$ is of the form $\MU x.\; \const{C}_i( \vec u )$.
\end{enumerate}
\end{definition}

@@

\(
\inferrule{
  x : \tau \in \FV( \Val )
  \quad
  (\MU x.\; t) \aequiv v
  \quad
  v \in \M( \tau )
  \quad
  v \not\in \mathcal{T}^\M_x(\Val)
}{
  \Val := \nf{(\Val \{ x \mapsto \MU x.\; t \})}
}
\)

In this update, we choose an unassigned variable in $\Val$
and assign it a value that is not $\alpha$-equivalent to one occurring in $\mathcal{T}^\M_x(\Val)$.
Since this update removes one variable from the set $FV( \Val )$ and does not add any additional variables to $\FV( \Val )$,
it can only be applied a finite number of times.
After assigning this value, we normalize all terms in the range of $\Val$.
%We then let $\Val^\ast \ec{x}$ be $\nf{(\Val' \ec{x})}$ for all $\ec{x}$.

@@

To show 3,
first note that since \rn{Unique} does not apply,
$\Val \ec{x} \aequiv \Val \ec{y}$ if and only if $\ec{x} = \ec{y}$.
Since \rem{TODO}
we have $\nf{(\Val \ec{x})} \aequiv \nf{(\Val \ec{y})}$ if and only if $\ec{x} = \ec{y}$.
We show that for any substitution $\sigma$ applied to $\Val$ while constructing its completion is such that
$\nf{(\Val \ec{x} \sigma)} \aequiv \nf{(\Val \ec{y} \sigma)}$ if and only if $\ec{x} = \ec{y}$.
\rem{TODO}
Let $P$ be the set of positions for which $\Val \ec{x}$ and $\Val \ec{y}$ are not $\alpha$-equivalent,
which is non-empty by assumption.

@@

If $t$ is a constructor term of the form $\const{C}( u_1, \ldots, u_n )$,
then $\M( t )$ is $\alpha$-equivalent with $\nf{(\MU x.\; \const{C}( \M( u_1 ), \ldots, \M( u_n ) ) )}$ for fresh $x$,
which by the induction hypothesis is $\alpha$-equivalent with $\nf{(\MU x.\; \const{C}( \Val^\ast \ec{ u_1 }, \ldots, \Val^\ast \ec{ u_n } ) )}$, call this term $t_u$.
By Lemma~\ref{lem:model-completion}.2 and due to our construction of $\Val^\ast$, we have $\Val^\ast \ec{t}$ is a term
$t_w$ of the form $\MU {\Varec{t}}.\; \const{C}( w_1, \ldots, w_n )$
where $t_w\interp{w_i} = \Val^\ast \ec{u_i}$ for $i = 1, \ldots, n$.
For each $i = 1, \ldots, n$, let $u_i'$ be the $i^{th}$ argument of $t_u$.
By Lemma~\ref{lem:mu-norm-arg}, $t_u\interp{u_i'} \vsim \Val^\ast \ec{u_i}$,
and thus $t_u\interp{u_i'} \vsim t_w\interp{ w_i }$. % for each $i = 1, \ldots, n$.
By Lemma~\ref{lem:mu-cong}, we have $t_u \vsim t_w$.
Since $\M( t ) \vsim t_u \vsim t_w = \Val^\ast \ec{t}$,
we have $\M( t ) \vsim \Val^\ast \ec{t}$.

>>>>

%\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\vec}[1]{\bar #1}
%\newcommand{\Ec}{\mathsf{E}}
\newcommand{\Ec}{E}
%\newcommand{\Ec}{\mathcal{E}}
%\newcommand{\Fc}{\mathsf{F}}
\newcommand{\Fc}{F}
%\newcommand{\Fc}{\mathcal{F}}
%\newcommand{\Gc}{\mathsf{G}}
%\newcommand{\Ac}{\mathsf{A}}
%\newcommand{\Dc}{\mathsf{D}}
%\newcommand{\Rc}{\mathsf{R}}
\newcommand{\vrange}{\mathsf{range}}
\newcommand{\vdom}{\mathsf{dom}}
\newcommand{\tEc}{\mathcal{T}(\Ec)}
\newcommand{\tcEc}{\mathcal{T}^\ast(\Ec)}
\newcommand{\rn}[1]{\textsf{\small #1}}
\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\tneq}{\not\teq}
\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
%\newcommand{\is}[1]{is\text{-} #1}
\newcommand{\is}[1]{\const{is#1}}
\newcommand{\ror}{\kern.45em \parallel \kern.45em}
\newcommand{\tpath}[2]{\mathcal{P}_{ #2 \rightarrow \_ }( #1 )}
\newcommand{\ttpath}[3]{\mathcal{P}_{ #2 \rightarrow #3 }( #1 )}
\newcommand{\interp}[1]{\llbracket #1 \rrbracket}
\newcommand{\dpath}[3]{\delta^{#1}_{#2}( #3 )}
\newcommand{\ec}[1]{[#1]}
\newcommand{\M}{\mathcal{M}}
\newcommand{\Val}{\mathcal{A}}
\newcommand{\Var}{\mathcal{V}}
%\newcommand{\Varec}[1]{\Var{\ec{#1}}}
\newcommand{\Varec}[1]{\widetilde{#1}}
\newcommand{\VAREC}[1]{\widetilde{\vphantom{\scriptstyle x^i}\smash{#1}}}
%\newcommand{\nf}[1]{{{#1}{\downarrow}}}
\newcommand{\nf}[1]{\lfloor#1\rfloor}
\newcommand{\aequiv}{\mathrel{=_\alpha}}
\newcommand{\vsim}{\aequiv}
\newcommand{\vsimv}[1]{\mathrel{=^{#1}_\alpha}}
\newcommand{\tpos}[2]{#1\!\mid_{#2}}
\newcommand{\muvar}{\mathcal{Var}}

@@

For any $\mu$-term $t$ of the form $\MU x.\; \const C( \vec u )$, we can
recursively construct the \emph{normal form} $\nf{t}$ of $t$,
that is a renaming of $v$ by replacing all of the bisimilar subterms of $t$ with $x$
%and by replacing all of the nonnormal subterms of $t$ with their corresponding normal forms.
and by normalizing $t$'s subterms.
For example, $\nf{\MU x.\; \const{C}( \MU y.\; \const{C}( x ) )} = \MU x.\; \const{C}( x )$.

\begin{lemma}
\label{lem:mu-norm-arg}
If $\vec u$ are normal and $t = \nf{\MU x.\; \const{C}( \vec u )} = \MU x.\; \const{C}( \vec w )$,
then $\vec u \vsim t\interp{\vec w}$.
\end{lemma}
\begin{proof}
%Since $u_i$ is normal, $w_i$ is the result of replacing all subterms $v$ of $u_i$ such that $t\interp{v} \aequiv t$ with $x$.
\rem{TODO}
\end{proof}

@@

\begin{definition}[Normal Model]
\afterDot%
\label{def:norm-model}%
\rm
A model $\M$ is \emph{normal} if and only if:
\begin{enumerate}
\item
for all types $\tau$,
$\M( \tau )$ is a maximal set of closed normal $\mu$-terms of that type that are
unique up to renaming of $\mu$-bound variables,
and acyclic when $\tau \not\in \Codata$,
\item
for all constructor terms $\const{C}( \vec t \,)$ of type $\tau$,
$\M( \const{C}( \vec t \,) )$ is the value
in $\M( \tau )$ that is $\alpha$-equivalent to
$\nf{\MU x.\; \const{C}( \M( \vec t \,) )}$ where $x$ is fresh, and
\item
for all selector terms $\const s^j_i( t )$ of type $\tau$,% for some $\vec u$,
$\M( \const s^j_i( t ) )$ is the value
in $\M( \tau )$ that is $\alpha$-equivalent to
$\M( t )\interp{u_j}$
when $\M( t )$ is of the form $\MU x.\; \const{C}_i( \vec u )$.
\end{enumerate}
\end{definition}

@@

\(
\inferrule{
  x : \tau \in \FV( \Val )
  \quad
  (\MU x.\; t) \aequiv v
  \quad
  v \in \M( \tau )
  \quad
  v \not\in \mathcal{T}^\M_x(\Val)
}{
  \Val := \nf{\Val \{ x \mapsto \MU x.\; t \}}
}
\)

In this update, we choose an unassigned variable in $\Val$
and assign it a value that is not $\alpha$-equivalent to one occurring in $\mathcal{T}^\M_x(\Val)$.
Since this update removes one variable from the set $FV( \Val )$ and does not add any additional variables to $\FV( \Val )$,
it can only be applied a finite number of times.
After assigning this value, we normalize all terms in the range of $\Val$.
%We then let $\Val^\ast \ec{x}$ be $\nf{\Val' \ec{x}}$ for all $\ec{x}$.

@@

To show 3,
first note that since \rn{Unique} does not apply,
$\Val \ec{x} \aequiv \Val \ec{y}$ if and only if $\ec{x} = \ec{y}$.
Since \rem{TODO}
we have $\nf{\Val \ec{x}} \aequiv \nf{\Val \ec{y}}$ if and only if $\ec{x} = \ec{y}$.
We show that for any substitution $\sigma$ applied to $\Val$ while constructing its completion is such that
$\nf{\Val \ec{x} \sigma} \aequiv \nf{\Val \ec{y} \sigma}$ if and only if $\ec{x} = \ec{y}$.
\rem{TODO}
Let $P$ be the set of positions for which $\Val \ec{x}$ and $\Val \ec{y}$ are not $\alpha$-equivalent,
which is non-empty by assumption.

@@

If $t$ is a constructor term of the form $\const{C}( u_1, \ldots, u_n )$,
then $\M( t )$ is $\alpha$-equivalent with $\nf{\MU x.\; \const{C}( \M( u_1 ), \ldots, \M( u_n ) ) }$ for some fresh $x$,
which by the induction hypothesis is $\alpha$-equivalent with $\nf{\MU x.\; \const{C}( \Val^\ast \ec{ u_1 }, \ldots, \Val^\ast \ec{ u_n } )}$, call this term $t_u$.
By Lemma~\ref{lem:model-completion}.2 and due to our construction of $\Val^\ast$, we have $\Val^\ast \ec{t}$ is a term
$t_w$ of the form $\MU {\Varec{t}}.\; \const{C}( w_1, \ldots, w_n )$
where $t_w\interp{w_i} = \Val^\ast \ec{u_i}$ for $i = 1, \ldots, n$.
For each $i = 1, \ldots, n$, let $u_i'$ be the $i^{th}$ argument of $t_u$.
By Lemma~\ref{lem:mu-norm-arg}, $t_u\interp{u_i'} \vsim \Val^\ast \ec{u_i}$,
and thus $t_u\interp{u_i'} \vsim t_w\interp{ w_i }$. % for each $i = 1, \ldots, n$.
By Lemma~\ref{lem:mu-cong}, we have $t_u \vsim t_w$.
Since $\M( t ) \vsim t_u \vsim t_w = \Val^\ast \ec{t}$,
we have $\M( t ) \vsim \Val^\ast \ec{t}$.
